---
title: "Project 2 -- Impact of WEATHER EVENTS in the US over Time"
author: "Bill Tihen"
date: "Dec 21, 2014"
output:
  html_document:
    toc: true
    keep_md: true
  pdf_document:
    toc: true
    highlight: zenburn

--- 

**For the course: Reproducable Data -- Project 2 -- Due Dec 21, 2014**

## Synopsis

This study from 1950 until 2011 studies the impact of health and the cost damage caused by weather events.  The data is presented in a variety of ways, so that public officals across the US can make appropriate decision based on the weather events in the local area.

**The following list actionable observations that can be used by public officials for planning** 

* areas near the coast are have the highest costs and health effects per event (hurricanes, tropical storms, tsunamis and tidal events are the most expensive per event), thus proper planning needs to be done so that emergencies services can handle large infrequent influxes -- probably in nearby, but inland areas less affect by the coastal events.  
* areas affected by tornados have the highest death and injury rates (because of the many weather events with small impact), thus proper emergency medical for frequent, but small numbers of injuries need to be planned.  
* areas affected by high heat, droughts and floods can expect signifigant health events (both injuries and deaths) and damages (both crop and property).  It may be worth investigating if most heat and drought related injuries can be prevented.  
* the number of events per year is increasing considerably -- this may mean that emergency response staff may have less recovery time between events.
* damages per year has been increasing since 1990, however, the years with very high expenses has increased in both magnitude and frequency since 1990 too -- thus it seems emergency response budgets will need to be increased at a corresponding rate.
* damages per event may be on the increase since 1990, and the very damaging events (outliers), have increased -- thus a relatively large worse case fiscal plan (on top of the normal emergency response budget) may be prudent.

**See the results section to view the supporting data.**

## Processing

1. **Load helpful libraries**

```{r libraries}
suppressWarnings(suppressMessages( library(plyr) ))
suppressWarnings(suppressMessages( library(knitr) ))
suppressWarnings(suppressMessages( library(ggplot2) ))
suppressWarnings(suppressMessages( library(data.table) ))
```


2. **Get and read the file:**

```{r get_data, cache=TRUE}
# define the data path, filename and URL to be downloaded
data_path = "data"
download_file = "storm-data.csv.bz2"
download_file_path = file.path( data_path, download_file )
url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# create the data folfder if it doesnt exist
if (!(file.exists( data_path ))) {
  dir.create( data_path, showWarnings = FALSE )
} 

# download the file to the data directory if it is not there
if (!(file.exists( download_file_path ))) {
  download.file(url, download_file_path, method="curl")
}
# read the file
raw_data <- read.table(bzfile( download_file_path ), sep=",", head=TRUE)
```

3. **Check that event labels are consistent with the 40+ NOA Standard events** listed at: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

```{r event_types, cache=TRUE}
# load into a data table - as it is easy to work with and fast to load
events_dt <- data.table(raw_data)

# how many data entries
events_total <- nrow( events_dt)
events_total

# check the event_types -- do they match NOA definitions?
levels( events_dt$EVTYPE )
```

4. **Obviously there are many duplicates, abbreviations, misspellings and non-categorized events -- all this needs to be cleaned up.**

```{r level_clean_up, cache=TRUE}
# copy factors to a new character field
events_dt$event_char <- as.character( events_dt$EVTYPE )

# simplify and clean up the data -- the coding is not consistent with the standards listed by
# by NOA as defined in their data definitions found at:
# https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
# categorize all summary data as summary data
events_dt$event_char <- gsub(".*summary.*", "summary", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*month.*", "summary", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*week.*", "summary", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*year.*", "summary", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*record.*", "summary", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*county.*", "summary", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*southeast.*", "summary", events_dt$event_char, ignore.case=TRUE)

# categorize all "other events" (accidents, mild weather, urban events and
# unoffical weather events, etc) as other.  For a list of official events, 
# see:  https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
# for the purposes of this analysis -- similar events have been further grouped from 
# 42 defined events to 30 types of weather events.
events_dt$event_char <- gsub(".*hyp.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*drown.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*dam.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*acc.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*mishap.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*flag.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub("\\?", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*no .*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*none.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*mild.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*other.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*urban.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*wall.*", "other", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*vog.*", "other", events_dt$event_char, ignore.case=TRUE)
# ice related weather events (using freezing early catches freezing events - which result in ice)
events_dt$event_char <- gsub(".*freezing.*", "ice", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*ice.*", "ice", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*icy.*", "ice", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*glaze.*", "ice", events_dt$event_char, ignore.case=TRUE)
# astronomical tidal events
events_dt$event_char <- gsub(".*tide$", "tide", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*tide.*", "tide", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub("^tidal.*", "tide", events_dt$event_char, ignore.case=TRUE)
# tornado related events
events_dt$event_char <- gsub(".*torn.*", "tornado", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*funn.*", "tornado", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*spout.*", "tornado", events_dt$event_char, ignore.case=TRUE)
# thunderstorm related events (tstm - is a NOA abbreviation for thunderstorms)
events_dt$event_char <- gsub(".*thun.*", "thunderstorm", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*tund.*", "thunderstorm", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub("^tstm.*", "thunderstorm", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*tstm.*", "thunderstorm", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*gustn.*", "thunderstorm", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*metro.*", "thunderstorm", events_dt$event_char, ignore.case=TRUE)
# lightning realated weather events
events_dt$event_char <- gsub(".*light.*", "lightning", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*ligtn.*", "lightning", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*lign.*", "lightning", events_dt$event_char, ignore.case=TRUE)
# blizzards
events_dt$event_char <- gsub(".*blizzard.*", "blizzard", events_dt$event_char, ignore.case=TRUE)
# hail
events_dt$event_char <- gsub(".*hail.*", "hail", events_dt$event_char, ignore.case=TRUE)
# sleet 
events_dt$event_char <- gsub(".*sleet.*", "sleet", events_dt$event_char, ignore.case=TRUE)
# avalanches
events_dt$event_char <- gsub("avalanc.*", "avalanche", events_dt$event_char, ignore.case=TRUE)
# costal flooding
events_dt$event_char <- gsub(".*coast.*", "coastal flood", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*cstl.*", "coastal flood", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*beach.*", "coastal flood", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*surge.*", "coastal flood", events_dt$event_char, ignore.case=TRUE)
# cold and wind chill events
events_dt$event_char <- gsub(".*cool.*", "cold", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*cold.*", "cold", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*chill.*", "cold", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*low.*", "cold", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*snow.*", "snow", events_dt$event_char, ignore.case=TRUE)
# fog related weather events
events_dt$event_char <- gsub(".*fog.*", "fog", events_dt$event_char, ignore.case=TRUE)
# dust storm and dust devil weather events
events_dt$event_char <- gsub(".*dust.*", "dust", events_dt$event_char, ignore.case=TRUE)
# dry weather (and droughts)
events_dt$event_char <- gsub(".*dry.*", "drought", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*excessive.*", "drought", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*drought.*", "drought", events_dt$event_char, ignore.case=TRUE)
# heat related weather events
events_dt$event_char <- gsub(".*heat.*", "heat", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*hot.*", "heat", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*warm.*", "heat", events_dt$event_char, ignore.case=TRUE)
# frost related weather events
events_dt$event_char <- gsub(".*frost.*", "frost", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*freeze.*", "frost", events_dt$event_char, ignore.case=TRUE)
# winter storms 
events_dt$event_char <- gsub(".*wint.*", "winter", events_dt$event_char, ignore.case=TRUE)
# fire related events
events_dt$event_char <- gsub(".*smoke.*", "fire", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*fire.*", "fire", events_dt$event_char, ignore.case=TRUE)
# rain related events
events_dt$event_char <- gsub(".*rain.*", "rain", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*mix.*", "rain", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*shower.*", "rain", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*precip.*", "rain", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*wet.*", "rain", events_dt$event_char, ignore.case=TRUE)
# hurricane events
events_dt$event_char <- gsub(".*hur.*", "hurricane", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*typ.*", "hurricane", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*floyd.*", "hurricane", events_dt$event_char, ignore.case=TRUE)
# rip current events
events_dt$event_char <- gsub(".*rip.*", "rip current", events_dt$event_char, ignore.case=TRUE)
# tropicl depressions
events_dt$event_char <- gsub(".*depress.*", "tropical depression", events_dt$event_char, ignore.case=TRUE)
# tropical storms
events_dt$event_char <- gsub(".*trop.*", "tropical storm", events_dt$event_char, ignore.case=TRUE)
# wind events (not part of a thunderstorm)
events_dt$event_char <- gsub(".*wind.*", "wind", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*turb.*", "wind", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*micro.*", "wind", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*wnd.*", "wind", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*burst.*", "wind", events_dt$event_char, ignore.case=TRUE)
# flooding events
events_dt$event_char <- gsub(".*flood.*", "flood", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*flooo.*", "flood", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*stream.*", "flood", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*water.*", "flood", events_dt$event_char, ignore.case=TRUE)
# land, mud and other debris slides
events_dt$event_char <- gsub(".*mud.*", "debris slide", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*rock.*", "debris slide", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*land.*", "debris slide", events_dt$event_char, ignore.case=TRUE)
# tsunami
events_dt$event_char <- gsub(".*tsunami.*", "tsunami", events_dt$event_char, ignore.case=TRUE)
# volcano events
events_dt$event_char <- gsub(".*volc.*", "volcano", events_dt$event_char, ignore.case=TRUE)
# various wave events at sea (marine waves)
events_dt$event_char <- gsub(".*surf.*", "marine waves", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*sea.*", "marine waves", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*sea.*", "marine waves", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*swell.*", "marine waves", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*wave.*", "marine waves", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*high.*", "marine waves", events_dt$event_char, ignore.case=TRUE)
events_dt$event_char <- gsub(".*seiche.*", "marine waves", events_dt$event_char, ignore.case=TRUE)

events_dt$event_type <- as.factor( events_dt$event_char )
levels( events_dt$event_type )
```

now that looks reasonable -- it's slightly more simplified than NOA, it groups many related events such as rainfall events, fog events, marine events, etc.

5. **Now that the levels are simplified, it will be helpful to make the damage amounts numeric (currently there is a unit and a character exponent, "K" for thousand, "M" for million and "B" for billion)**

```{r calculate_cost, cache=TRUE}
                                     
# convert the crop damage into us dollars
events_dt$crop_mult <- list(0)
events_dt$crop_mult[ events_dt$CROPDMGEXP == "K" ] <- 1000
events_dt$crop_mult[ events_dt$CROPDMGEXP == "M" ] <- 1000000
events_dt$crop_mult[ events_dt$CROPDMGEXP == "B" ] <- 1000000000
events_dt$crop_usd <- events_dt$CROPDMG * events_dt$crop_mult

# convert the property damage into us dollars
events_dt$property_mult <- list(0)
events_dt$property_mult[ events_dt$PROPDMGEXP == "K" ] <- 1000
events_dt$property_mult[ events_dt$PROPDMGEXP == "M" ] <- 1000000
events_dt$property_mult[ events_dt$PROPDMGEXP == "B" ] <- 1000000000
events_dt$property_usd <- events_dt$PROPDMG * events_dt$property_mult
```

6. **It will probably also be useful to know the total damages of both crop and property damages*

```{r combined_damage, cache=TRUE}
# calculate combined damage costs
events_dt$total_usd  <- events_dt$crop_usd + events_dt$property_usd
```

7. **It will probably also be useful to know the combined deaths and injurs in an event**

```{r combined_health, cache=TRUE}
# calculate combined health effects (death and injuries)
events_dt$health_count <- events_dt$FATALITIES + events_dt$INJURIES
```

8. **It is important to know over what time frame this data has been collected**

```{r time_frame, cache=TRUE}
# convert the begin data to extract just the year as a numeric value
#events_dt$year <- as.numeric(format( as.Date(as.character(events$BGN_DATE), %m/%d/%Y %T), "%Y""))
events_dt$date <-  as.Date( as.character(events_dt$BGN_DATE), format = " %m/%d/%Y %T" )
events_dt$year <- as.numeric( format( events_dt$date, "%Y" ) )

# find the most recent event year 
max_year   = max( events_dt$year )
max_year
# find the oldest event year
min_year   = min( events_dt$year )
min_year
# calculate the study time span
years_span = max_year - min_year
years_span
```

9. **Check the frequency of weather events -- is it similar over time? (are yearly averages meaningful?)

```{r event_consistency, cache=TRUE}
# understand frequency by year
events_in_year <- as.data.frame(table(events_dt$year))
# provide useful column names
colnames(events_in_year)[1] <- "year"
colnames(events_in_year)[2] <- "events_count"
events_in_year$year <- as.numeric( as.character( events_in_year$year ) )

# compare events_in_year to number of events in the original data 
summed_by_year <- sum( events_in_year$events_count )
summed_by_year
events_total
```

10. **Calculate the number of occurances for each type of event**

```{r event_count, cache=TRUE}
# count each type of event
event_count <- as.data.frame(table(events_dt$event_type))
# provide useful column names
colnames(event_count)[1] <- "event_type"
colnames(event_count)[2] <- "events_count"

# compare events_in_year to number of events in the original data 
summed_by_type <- sum( events_in_year$events_count )
summed_by_type
events_total
```

11. **Summarize effects (HEALTH effects: deaths and injuries and DAMAGE costs: crop and property) by the simplified event types (events_dt$event_type)**

```{r summarize_by_event, cache=TRUE}
# create a summary data table of each event type with important effects summed
effects <- events_dt[, lapply(.SD, sum, na.rm=TRUE), by="event_type", .SDcols=c("FATALITIES", "INJURIES", "property_usd", "crop_usd", "health_count", "total_usd") ]
```

12. **Calculate effect per event (intensity has not been considered in this study -- which is likely proportional to cost per event)**

merge the effect datatable with the number of events per event category:

```{r effect_per_event, cache=TRUE}
# merge the effect summary with the event_count datatable
effects <- merge(effects, event_count, by="event_type")

# calculate the average effects per event
effects$avg_health_per_event <- round( effects$health_count / effects$events_count, 2 )
effects$avg_cost_per_event   <- round( effects$total_usd    / effects$events_count, 2 )
```

13. **Check if cost per event (probably related to event intensity -- once inflation is subtracted) is increasing -- but it can be a good way to estimate the near future expectations**

```{r cost_year, cache=TRUE}
# create a summary data table of each event type with important effects summed
costs <- events_dt[, lapply(.SD, sum, na.rm=TRUE), by="year", .SDcols=c("FATALITIES", "INJURIES", "property_usd", "crop_usd", "health_count", "total_usd") ]
```

14. **Calculate yearly effects**

```{r costs_per_year, cache=TRUE}
# merge the effect summary with the event_count datatable
costs <- merge(costs, events_in_year, by="year")

# calculate the average effects per event
costs$avg_health_per_event <- round( costs$health_count / costs$events_count, 2 )
costs$avg_cost_per_event   <- round( costs$total_usd    / costs$events_count, 2 )
```

## RESULTS

1. This study spans from **`r min_year`** to **`r max_year`**, a time span of **`r years_span`**.

2. **The number of events each year is increasing strongly:**

```{r events_per_year, cache=TRUE}
#plot( events_in_year$year, events_in_year$events_count, 
#      main="Weather Events over Time", 
#      xlab="Year", ylab="Number of Events", pch=19)

qplot(year, events_count, data=events_in_year, geom=c("point", "smooth"), 
      main="Weather Events over Time", 
      xlab="Year", ylab="Number of Events")
```

3. **The costs per event has possibly started to rise since 1990 (however, outliers have been increasing in both magnitude and frequency):**

```{r costs_per_events_plot, cache=TRUE}
#plot( costs$year, costs$avg_cost_per_event, 
#      main="Cost per Event over Time", 
#      xlab="Year", ylab="Cost per Events", pch=19)

qplot(year, avg_cost_per_event, 
      data=costs, geom=c("point", "smooth"), 
      #method="lm", formula=y~x, 
      main="Cost per Event over Time", 
      xlab="Year", ylab="Cost per Events")
```

4. **The costs per year has started to rise since 1990 and the variation (& outlies may be increasing):**.**

```{r costs_per_year_plot, cache=TRUE}
#plot( costs$year, costs$total_usd, 
#      main="Cost per Year over Time", 
#      xlab="Year", ylab="US Dollars", pch=19)

qplot(year, total_usd, 
      data=costs, geom=c("point", "smooth"), 
      #method="lm", formula=y~x, 
      main="Cost per Year over Time", 
      xlab="Year", ylab="US Dollars")

```

More investigation would be needed, this may be due to proper preventative measures.

5. **The most common events that affect health (sorted by deaths) are shown in the following table:**

```{r health_sorts, cache=TRUE}
# copy the orginal data table so that removing columns doesn't affect the original
health <- as.data.frame(effects)
health <- data.table(health)
# remove costs -- and focus only on health impacts
health[ , c("property_usd", "crop_usd", "total_usd", "avg_cost_per_event"):=NULL ]

# sorting by deaths
health_by_deaths <- health[ order( -health$FATALITIES) ]
kable( head( health_by_deaths ) )
# sorting by injuries
health_by_injuries <- health[ order( -health$INJURIES) ]
kable( head( health_by_injuries ) )
# sorting by all health impacts
health_per_event <- health[ order( -health$avg_health_per_event) ]
kable( head( health_per_event ) )
```

6. **event types sorted by costs**

```{r cost_sorts, cache=TRUE}
# copy the orginal data table so that removing columns doesn't affect the original
cost <- as.data.frame(effects)
cost <- data.table(cost)
# remove the health data -- just view costs
cost[ , c("FATALITIES", "INJURIES", "health_count", "avg_health_per_event"):=NULL ] 

# sort by crop damage impact
cost_by_crop <- cost[ order( -cost$crop_usd) ]
kable( head( cost_by_crop ) )
# sort by property damage impact
cost_by_property <- cost[ order( -cost$property_usd) ]
kable( head( cost_by_property ) )
# sort by costs per event
cost_per_event <- cost[ order( -cost$avg_cost_per_event) ]
kable( head( cost_per_event ) )
```


## Analysis Caveots

* The fiscal analysis has ignored the effect of inflation.
* Weather intensity changes over time have not been investigated.


## Discussion, Questions and Surprises

The data is summarized in the synopsis.

Here I will only mention surprises:

* droughts and heat were in the top 5 for health impacts -- for both injuries and deaths -- can these be prevented?
* although weather event frequency has increased, the cost per year has not increased at a corresponding rate -- have prevention methods been extreemly effective?
* does the increased variation in cost per year indicate an increase in the likelihoot of more intense weather events?
